STAT 450 Project: Real Estate

### 2020-03-20 src/2nd_draft.Rmd
grammar change, fully revised, added conclusion

Our client provides us the past 5 years' property assessment data in BC. Since we are only interested in predicting mill rate for metro Vancouver and specific class code, we get a subset of properties that satisfy our interest:

@@

Moreover, we select 5 features that could be relevant to mill rate, which are:

@@

To reduce the dimension of our data, we aggregate all properties in the same region, tax class code, and year are aggregate into a group, becasue these properties have same mill rate,  which is our response variable. Here is the summary statistics for these groups:

- Mill rate 
- Total Assessment 
- Total Land Assessment 
- Total Improvement Assessment 
- Total number of properties 
- Taxclasscode
- Municipality
- Year

@@

Before any prediction on the future mill rate of Metro Vancouver’s real estate market was made, exploratory data analysis was performed to explore and visualize the main characteristics of our dataset and found relationships between Mill Rate vs. Assessment, Mill Rate vs. Land Total, and Mill Rate vs. Improvement Total were performed. From our initial analysis, outliers in municipalities had been found. Data transformation - calculating the average total assessment, was used to reduce the effect of outliers.

Mill rate is mainly affected by assessment, so scatter plots of mill rate vs. total, land, and improvement assessment were created to see the correlation between each pair of the two factors. Kruskal Wallis analysis was also performed to see the correlation between mill rate vs. tax class and mill rate vs. municipality.

@@

In this study, we built linear models and evaluated their performances by goodness of fit and prediction power, that is, how well the model explains the data and how well it can predict future values. The definition of goodness of fit and prediction power is given below.

 - **Goodness of fit** is defined as the extent to which the sample data are consistent with the model, which examines how well the model explains the data. R squared and adjusted R squared are the most well known measures of goodness of fit and higher R squared and adjusted R squares indicate better goodness of fit.

 - **Prediction power** measures how well models can predict the future values. We use mean squared prediction power to compare the prediction performance across all of our fitted linear predictive models. To do that, we will divide the data set into training data - used to build our models, and testing data - used to evaluate the prediction power of our models.

@@

TaxClassCode, Municipalities, Assess Total, Land Assessment Total, Improvement Assessment total, Number of properties are considered. The full linear model (ORL full) using all the available features was built first. Based on the results of EDA, we selected a list of significant variables and included them in another linear model (OLR transformed). To compare the effect of linear models with and without features, a null model (OLR null) with no features used, was also constructed.

@@

Other than ordinary linear regressions, we are also interested in the performance of more advanced linear regressions like Ridge and Lasso. Ridge and Lasso have a different objective function to optimize; they take penalty in sum of absolute values and sum of squared absolute values of weights respectively. These models have an interesting characteristic, that is, a weight is assigned to each feature. This might be relevant to our study as mill rate can be affected by each feature to a different extent. Also, another advantage of these model is that MSPE is more stable as the variance of MSPE is reduced.

@@

To examine the goodness of fit and prediciton power, a 50-run of 10-fold cross validation would be performed in this study. For each run, a 10-fold corss validation was used to train the five models and make predictions on training and testing data respectively. Then, the MSEs calculated from the training data and MSPEs calculated from the testing data were stored in vectors of corresponding mdoels.

After the 50 runs, a vector of MSEs and a vectors of MSPEs for each model were therefore constructed successfully. Based on these vectors, boxplots would be used to show the mean and spread of MSE and MSPE aross all models respectively.

@@

Since mill rate is mainly affected by total assessment, which is sum of total land assessment and total improvement assessment, scatter plots between mill rate and all kinds of assessment among different municipalities are created to show the relationship in Figure 1, Figure 2, Figure 3

@@

There is no clear trend between mill rate and any kind of assessment. Plots also have shown that most points are condensed on the left horizontal axis, since some municipalities have large assessment.

To reduce the effect of large assessment of some municipalities, total assessment across municipalities are transformed by taking total assessment dividing by number of properties of each municipality and tax class. The transformed data is named **"Average Total Assessment"**. A scatter plot between mill rate and average total assessment is showed in Figure 4.

@@

Plot from Figure 4 has shown that mill rate tends to decrease as average total assessment increase. Also they have moderately strong linear correlation.

@@

Figure 5 has shown that most municipalities have different mean and variance in terms of mill rates.

@@

A statistical test called Kruskal-Wallis Test is also performed to test their distribution. It is used to decide if population distributions are identical, and the corresponding p-value which is smaller than 0.05 can indicate that the data has nonidentical distributions.

@@

The p-values in Table 1 supports that there is nonidentical distributions of mill rates across municipalities and tax classes.

@@

Below is a comparison of the five models (Lasso, Ridge, OLR transformed, OLR full and OLR null) as for goodness of fit and predcition power. Goodness of fit is measured by MSE of the training data, whereas prediction power is measured by MSPE of the testing data. Generally, smaller MSEs and MSPEs indicate better fit and prediction power respectively.

The distributions of goodness of fit and prediction power across models are displyed in Figure 7 and Figure 8, respectively, the distribution of null model is removed since it has larger values compared to all other models.

@@

boxplot(mspe1.la, 
        mspe1.ri, 
        mspe1.reduced, 
        mspe1.f, 
        names = c("LASSO", 
                  "Ridge",
                  "Transformed", 
                  "Full"
                  ), 
        col = c("steelblue","gray80",
                "pink", 
                "tomato"), 
        cex.axis = 1, 
        cex.lab = 1, 
        cex.main = 2,
        main = " Figure 7: Goodness of Fit",
        ylab = expression(hat(MSE)))

boxplot(mspe.la, 
        mspe.ri,
        mspe.reduced, 
        mspe.f,
        names = c("LASSO", 
                  "Ridge", 
                  "Transformed", 
                  "Full"
                  ), 
        col = c("steelblue","gray80", 
                "pink",
                "tomato"), 
        cex.axis = 1, 
        cex.lab = 1, 
        cex.main = 2,
        main = "Figure 8: Prediction Power",
        ylab = expression(hat(MSPE)))
```

@@

As for goodness of fit, Lasso, Ridge and OLR reduced have a similar performance; MSEs are around 4.6, but OLR transformed perform slightly better and has a smaller spread (variance) of MSE. OLR full performs worse than the the three, around 5.4, and OLR null has the greatest MSE, around 35.

Similarly, as for prediction power, Lasso, Ridge and OLR transformed perform roughly the same; MSPEs are close to 6. OLR full is worse with MSPE around 6.9. OLR Null performs the worst with MSPE over 35.

Therefore, we conclude that from the results of a 50-run of 10-fold cross validation, Lasso, Ridge and OLR transformed have the best goodness of fit and prediction power. OLR null perfroms much worse than all the other four models.

For a detailed look of how the models were fitted and evaluated, please refer to the **Appendix**.

@@

##_____________________________
## TODO: delete pct change part, add prediction based on our chosen model (Peter)
## Conclusion

From our exploratory data analysis, we can see that assessment total, land total, tax code, year and municipalities are all significant in our model. The client suggested that we can transform numerical factors such as assessment total into percentage change. This data transformation does not perform as well as we expected, which only yields an R^2 of around 0.2 across all of our fitted models. We believe the method that the client has suggested might leave out some important information about the housing market in each municipality. We have also found that when comparing the correlation between mill rate against features in our models, Vancouver is an outlier. We would like to further investigate Vancouver as a special case. We are also interested in the effect of different tax classes in predicting mill rate. Further analysis and prediction will be performed based on our hypothesis.

All fitted linear model were able to make accurate prediction based on means squared error, but there is still a very high chance that our model is overfitted. For our next report, we are going to use cross-validation to reduce the effect of overfitting. 
##_______________________________

>>>>

Our client provides us the past 5 years' property assessment data in BC. Since the only interest is in predicting mill rate for metro Vancouver and specific class code, a subset of properties that satisfy our interest have been selected:

@@

Moreover, 5 features have been selected that could be relevant to the mill rate:

@@

To reduce the dimension of our data, all properties in the same region, tax class code, and year are aggregated into a group because these properties have the same mill rate,  which is our response variable. Here is the summary statistics for these groups:

- Mill rate ("rate": 2016-2019)
- Total Assessment ("assessTotal": 2016-2019)
- Total Land Assessment ("landTotal": 2016 - 2019)
- Total Improvement Assessment ("improvementTotal": 2016 - 2019)
- Total number of properties ("propertyCount")
- Taxclasscode ("TaxClassCode": 01, 05, 06)
- Municipality ("AddressAssessorMunicipalityDesc")
- Year

@@

Before any prediction on the future mill rate of Metro Vancouver’s real estate market was made, exploratory data analysis was performed to explore and visualize the main characteristics of our dataset. Correlation analyses between Mill Rate vs. Assessment, Mill Rate vs. Land Total, and Mill Rate vs. Improvement Total were performed. From our initial analysis, outliers in municipalities had been found. Data transformation - calculating the average total assessment, was used to reduce the effect of outliers.

Mill rate was mainly affected by assessment, so scatter plots of mill rate vs. total, land, and improvement assessment were created to see the correlation between each pair of the two factors. Kruskal Wallis analysis was also performed to see the correlation between mill rate vs. tax class and mill rate vs. municipality.

@@

In this study,  linear models were built and the performance of each model was evaluated by the goodness of fit and prediction power, that is, how well the model explains the data and how well it can predict future values. The definition of the goodness of fit and prediction power is given below.

 - **Goodness of fit** is defined as the extent to which the sample data are consistent with the model, which examines how well the model explains the data. R-squared and adjusted R-squared are the most well-known measures of goodness of fit and higher R-squared and adjusted R squares indicate better goodness of fit.

 - **Prediction power** measures how well models can predict future values. Mean squared prediction power was used to compare the prediction performance across all of our fitted linear predictive models. To do that, the data set was divided into training data - used to build our models, and testing data - used to evaluate the prediction power of our models.

@@

The full linear model (ORL full) was built first. TaxClassCode, Municipalities, Assess Total, Land Assessment Total, Improvement Assessment total, Number of properties were considered in this model. Based on the results of EDA, we selected a list of significant variables and included them in another linear model (OLR transformed). To compare the effect of linear models with and without features, a null model (OLR null) with no features used, was also constructed.

@@

Other than ordinary linear regressions, we are also interested in the performance of more advanced linear regressions like Ridge and Lasso. Ridge and Lasso have a different objective function to optimize; they take a penalty in the sum of absolute values and sum of squared absolute values of weights respectively. These models have an interesting characteristic, that is, a weight is assigned to each feature. This might be relevant to our study as the mill rate can be affected by each feature to a different extent. Also, another advantage of these models is that MSPE is more stable as the variance of MSPE is reduced.

@@

To examine the goodness of fit and prediction power, a 50-run of 10-fold cross-validation was performed in this study. For each run, 10-fold cross-validation was used to train the five models and make predictions on training and testing data respectively. Then, the MSEs calculated from the training data and MSPEs calculated from the testing data were stored in vectors of corresponding models.

After the 50 runs, a vector of MSEs and a vector of MSPEs for each model were therefore constructed successfully. Based on these vectors, side-by-side boxplots were used to show the mean and the spread of MSE and MSPE across all models respectively.

@@

Since the mill rate is mainly affected by the total assessment, which is the sum of total land assessment and total improvement assessment, scatter plots between mill rate and all kinds of assessment among different municipalities are created to show the relationship in Figure 1, Figure 2, Figure 3

@@

There is no clear trend between the mill rate and any kind of assessment. Plots have shown that most points are condensed on the left horizontal axis since some municipalities have larger assessment values than others.

To reduce the effect of large assessment in some municipalities, total assessment across all municipalities is transformed by taking total assessment dividing by the number of properties of each municipality and tax class. The transformed data is named **"Average Total Assessment"**. A scatter plot between mill rates and average total assessments is shown in Figure 4.

@@

The plot from Figure 4 has shown that the mill rate tends to decrease as the average total assessment increase. Also, they have a moderately strong linear correlation.

@@

Figure 5 shows that most municipalities have different mean and variance in terms of mill rates.

@@

A statistical test called the Kruskal-Wallis Test was performed to test their distribution. It was used to decide if population distributions were identical, and the corresponding p-value which is smaller than 0.05 indicated that the data have nonidentical distributions.

@@

The p-values in Table 1 support that there is nonidentical distributions of mill rates across municipalities and tax classes.

@@

Below is a comparison of the five models (Lasso, Ridge, OLR transformed, OLR full and OLR null) using the goodness of fit and prediction power. The goodness of fit was measured by MSE of the training data, whereas prediction power was measured by MSPE of the testing data. Generally, smaller MSEs and MSPEs indicate better fit and prediction power respectively.

The distributions of the goodness of fit and prediction power across models are displayed in Figure 7 and Figure 8, respectively, the distribution of null model is removed since it has larger values compared to all other models.

@@

boxplot(mspe1.la, 
        mspe1.ri, 
        mspe1.reduced, 
        mspe1.f, 
        names = c("LASSO", 
                  "Ridge",
                  "Transformed", 
                  "Full"
                  ), 
        col = c("steelblue","gray80",
                "pink", 
                "tomato"), 
        cex.axis = 1, 
        cex.lab = 1, 
        cex.main = 1,
        main = " Figure 7: Goodness of Fit",
        ylab = expression(hat(MSE)))

boxplot(mspe.la, 
        mspe.ri,
        mspe.reduced, 
        mspe.f,
        names = c("LASSO", 
                  "Ridge", 
                  "Transformed", 
                  "Full"
                  ), 
        col = c("steelblue","gray80", 
                "pink",
                "tomato"), 
        cex.axis = 1, 
        cex.lab = 1, 
        cex.main = 1,
        main = "Figure 8: Prediction Power",
        ylab = expression(hat(MSPE)))
```

@@

As for goodness of fit, Lasso, Ridge, and OLR transformed have similar performance; MSEs across these models were around 4.6, but OLR transformed performed slightly better and had a smaller spread (variance) of MSE. OLR full performed worse than the other three, around 5.4, and OLR null had the greatest MSE, around 35.

Similarly, as for prediction power, Lasso, Ridge, and OLR transformed performed roughly the same; MSPEs across these models were close to 6. OLR full was worse with MSPE around 6.9. OLR Null performed the worst with MSPE over 35.

Therefore, we conclude that from the results of a 50-run of 10-fold cross-validation, Lasso, Ridge, and OLR transformed had the best goodness of fit and prediction power. OLR null perfromed much worse than all the other four models.

For a detailed look at how the models were fitted and evaluated, please refer to the **Appendix**.

@@

## Conclusion

From the exploratory data analysis, we have found that Total Assessments contain outliers in major municipalities such as Vancouver and Burnaby. In order to reduce the effect of outliers, the Total Assessments were transformed by taking their average over the number of properties in each municipality. The correlation analyses had shown that the transformed assessment total is the only continuous variable that has a relatively strong correlation with the mill rate. The Kruskal-Wallis had suggested that mill rates in each municipality and tax class are significantly different. From these results, we have decided to use Average Assessment Total, TaxClassCode and Municipality to fit our reduced model.

Transformed OLR, Ridge Regression, and LASSO were able to make good predictions based on mean squared error and mean squared prediction error from cross-validation. Since the client prefers a simpler model, we choose the transformed model to make our 2020 prediction.  
##_______________________________
